---
title: "belong_stub2"
author: "GUID"
date: "2023-06-07"
output: html_document
---

You'll see that the code chunks that have the wrong code in them have `eval = FALSE` at the top. This means that when you knit the file, it won't evaluate the code in this chunks - this means you can still knit the file (as it won't knit if there are errors)

## Activity 1

Click Session then Restart R.

```{r}
rm(list = ls())
```


## Activity 2

```{r chunk2a, eval=FALSE}
# this is the error mode code
demographic_data <- read_csv("demographic_data.csv")
questionnaire_data <- read_csv("questionnaire_data.csv")
```

```{r chunk2b}
# add the correct code in this chunk


```

What was the error and how did you fix it?


## Activity 3

```{r chunk3a, eval=FALSE}
# this is the error mode code
full_dat <- inner_join(x = demographic_data,
                       y = questionnaire_data, 
                       by = "participant", "level")
```

```{r chunk3b}
# add the correct code in this chunk


```

What was the error and how did you fix it?

## Activity 4

```{r chunk4a, eval=FALSE}
# this is the error mode code
first_year <- full_dat %>%
  filter(level = "First year")
```

```{r chunk4b}
# add the correct code in this chunk


```

What was the error and how did you fix it?

## Activity 5

```{r chunk5a, eval=FALSE}
# this is the error mode code
first_year <- first_year %>%
  mutate(gender_coded = factor(gender, levels = c(1,2,3),
                         labels = c("Man", "Woman", "Non-binary")))

first_year %>%
  count(gender_coded)
```

```{r chunk5b}
# add the correct code in this chunk


```

What was the error and how did you fix it?


## Activity 6

```{r chunk6a, eval=FALSE}
# this is the error mode code
first_year <- first_year %>%
  selct(Participant, 
         "gender" = gender_coded,
         age,
         belongingness:self_confidence) 
```

```{r chunk6b}
# add the correct code in this chunk


```

What was the error and how did you fix it?


## Activity7 

```{r chunk7a, eval=FALSE}
# this is the error mode code
scores_gender <- first_year %>%
  group_by(participant, gender) %>%
  summarise(belong = mean(belongingness),
            engagement = mean(engagement),
            confidence = mean(self_confidence))
```

```{r chunk7b}
# add the correct code in this chunk


```

What was the error and how did you fix it?


## Activity 8

```{r chunk8a, eval=FALSE}
# this is the error mode code
ggplot(first_year, aes(x = gender, y = belongingness) +
  geom_violin() +
  geom_boxplot(width = .2) 
```

```{r chunk8b}
# add the correct code in this chunk


```

What was the error and how did you fix it?



## Activity 9

```{r chunk9a, eval=FALSE}
# this is the error mode code
ggplot(first_year, aes(x = belongingness, y = engagement,fill = gender)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_colour_viridis_d(option = "B")
```

```{r chunk9b}
# add the correct code in this chunk


```

What was the error and how did you fix it?



